---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


# -----------------------------------------------------------------------
# 
##Getting started
```{r starting, warning=FALSE, message=FALSE}
library(dplyr)
rm(list=ls()) #Cleans previous variables
knitr::opts_chunk$set(results="hide", warning=FALSE, message=FALSE) #environment variables
```

## Loading and preprocessing the data
```{r loading_data}
#Reads data
data=read.csv("activity.csv", header=TRUE)
data$date=as.Date(data$date) # Formats date
```

## What is mean total number of steps taken per day?
```{r Question_1}
#Creates different dataframe with no NAs
data.na=na.omit(data)

#Creates table with total steps by day
summary_day=data.na %>%
    select(1:2) %>% #drops the interval column
        group_by(date) %>%
            summarise_each(funs(sum))

#Mean and median total number of steps taken per day
mean=mean(summary_day$steps)
median=median(summary_day$steps)

#Plots histogram of steps per day
hist(summary_day$steps, breaks=20, main="Histogram of total steps per day", xlab = "Total steps", col="green")

```

###Solution:
####The mean number of steps per day is `r format(round(mean, 1), nsmall = 1)` and the median is `r format(round(median, 1), nsmall = 1)`

------------------------------------------------------
------------------------------------------------------
------------------------------------------------------

## What is the average daily activity pattern?
```{r Question_2}
#Creates table with average steps per interval
summary_interval=data.na %>%
    select(c(1,3)) %>% #drops the dates column
        group_by(interval) %>%
            summarise_each(funs(mean))
#Max number of steps
max=max(summary_interval$steps)
#Interval for where the maximum number of steps occur
inter=summary_interval$interval[which(summary_interval$steps == max(summary_interval$steps))]

plot(summary_interval, main="Steps as a function of day interval", xlab = "Interval", ylab = "Steps", type="l",lwd=2)
```

###Solution:
####The maximum number of steps is `r format(round(max, 1), nsmall = 1)` and occurs during the interval `r inter`

------------------------------------------------------
------------------------------------------------------
------------------------------------------------------

## Imputing missing values
```{r Question_3}
#Number of rows where steps=NA
number.na.steps=sum(is.na(data$steps))

data_NA=subset(data, is.na(data$steps))

#Generate new dataframe where NA values are replaced by average value for the corresponding interval
data_replace=data
for(i in 1:nrow(data)){
    if(is.na(data[i,1])) { # if step value is NA
        interval=data$interval[i] # get what interval corresponds to that NA value
        data_replace[i,1]=summary_interval[which(summary_interval$interval == interval),2]
    }
}

#Creates table with the total steps by day
summary_day_rep=data_replace %>%
    select(1:2) %>%
        group_by(date) %>%
            summarise_each(funs(sum))
#Mean and median total number of steps taken per day
mean2=mean(summary_day_rep$steps)
median2=median(summary_day_rep$steps)

#Plots histogram of steps per day
hist(summary_day_rep$steps, breaks=20, main="Histogram of total steps per day after NA replacement",
     xlab = "Total steps", col="red")
```

###Solution:
####There are `r number.na.steps` rows that have a NA value. After replacing the NA values with the average number of steps for that particular interval, the mean number of steps per day is `r format(round(mean2, 1), nsmall = 1)` and the median is `r format(round(median2, 1), nsmall = 1)`. Very similar to the values before the replacement.

------------------------------------------------------
------------------------------------------------------
------------------------------------------------------


## Are there differences in activity patterns between weekdays and weekends?
```{r Question_4}
#Formats date as dates and adds column with weekday
data_replace$weekday=weekdays(data_replace$date) # Adds column with day

#Generate separate datasets for weekdays and weeekens only
data_replace.weekend=subset(data_replace, weekday=="Saturday"|weekday=="Sunday")
data_replace.weekday=subset(data_replace, !(weekday=="Saturday"|weekday=="Sunday"))

#Creates table with average steps per interval during weekdays
summary_interval_weekday=data_replace.weekday %>%
    select(c(1,3)) %>% #drops the dates column
        group_by(interval) %>%
            summarise_each(funs(mean))

#Creates table with average steps per interval during weekends
summary_interval_weekend=data_replace.weekend %>%
    select(c(1,3)) %>% #drops the dates column
        group_by(interval) %>%
            summarise_each(funs(mean))

weekday_walk=sum(summary_interval_weekday$steps) # Average total steps per day during weekdays
weekend_walk=sum(summary_interval_weekend$steps) # Average total steps per day during weekends

plot(summary_interval_weekday, col="blue", main="Walking activity", xlab = "Interval", ylab = "Steps", type="l",lwd=2)
points(summary_interval_weekend, col="red", xlab = "Interval", ylab = "Steps", type="l",lwd=2)
legend(1800,220, c("Weekdays", "Weekends"), lty=c(1,1), lwd=c(2,2), col=c("blue","red"))
```

###Solution:
####It appears that walking activity begins and ends earlier during the weekdays than in the weekends, it also looks like the subject walks less in the weekdays (an average of `r format(round(weekday_walk, 1), nsmall = 1)` steps per day) than in the weekends (with an average of `r format(round(weekend_walk, 1), nsmall = 1)` steps per day)
